Assessment of drought in continental Chile for 1981–2023 by climate variables of water supply and demand, soil moisture, and vegetation

### 2024-01-15 conference/IGARSS2024.qmd
avanzar en paper metodologia

---
title: "DDS4Chl: A dataset of drought indices of water demand and supply and its impact on vegetation for continental Chile"
format:
  ieee-pdf:
    keep-tex: false  
  ieee-html: default
title-block-banner: true
author:
  - id: dfolio
    name: Francisco Zambrano
    affiliations:
      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745
        orcid: 0000-0001-6896-8534
        email: francisco.zambrano@umayor.cl
        #url: https://dfolio.fr/
        attributes:
        corresponding: true
        #photo: david-folio.png
  - id: iduran
    name: Iongel Duran-Llacer
    affiliations: 
      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745

note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
#keywords: [IEEE, IEEEtran, journal, Quarto, Pandoc, template]
#funding: 
funding: 
  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022 and 11190360."
pageheader:
  left: IGARSS, January 2024
  right: 'D. Folio:  A Sample Article Using quarto-ieee'
bibliography: references.bib
date: 2024-01-05
---

@@

A persistent drought is impacting Chile. It affects the hydrological system and vegetation development. Research studies have focused on the central part of the country. This is due to a persistent period of water scarcity. This scarcity has been found to be a megadrought. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December. The SPI only considers precipitation as a drought indicator. It does not account for atmospheric evaporative demand (AED), soil moisture, or their combined vegetation productivity. We present a developing database (DDS4Chl) of drought indices for continental Chile since 1981. The indices measure water demand, supply and its impact on vegetation. We derived the SPI for water demand. We also derived the Evaporative Demand Drought Index (EDDI) for water supply. The Standardized Precipitation Evapotranspiration Index (SPEI) shows the combined effect of AED and precipitation. We estimate the standardized anomaly of cumulative soil moisture at one meter (zcSM) and the standardized anomaly of cumulative NDVI (zcNDVI) to show the impact of water demand and supply. Finally, we present the historical linear trend of the drought indices in continental Chile. Also we calculated the temporal correlation between the indices of water supply, water demand, and soil moisture with the zcNDVI (proxy of vegetation productivity). The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.
<br>

**Index Terms—** drought, Chile, water supply, water demand

@@

@@

Continental Chile has at least seven major climatic types. They range from low desert in the north to alpine tundra and glaciers in the east and southeast. The south has an Oceanic climate and central Chile has a Mediterranean climate [@kottek2006]. For the analysis, continental Chile was also divided into five zones according to a latitudinal gradient: "Norte Grande", "Norte Chico",  "Zona Central", "Zona Sur", and "Zona Austral". (@fig-studyArea)

```{r fig-studyArea}
#| fig-pos: "!ht"
#| fig-cap: "(a) Chile and the five zones 'norte grande', 'norte chico',  'zona central', 'zona sur', and 'zona austral'. (b) Topography reference map. (c) Land cover classes for 2021. (d) Persistent land cover classes (> 80%) for 2001-2021"
#| out-width: "\\textwidth"

@@

Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var). Indicating either a reduction of vegetation surface, biomass, or status. It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."

```{r fig-zcNDVI_var, out.extra='style="float:right; padding:10px"'}
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
```

Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation. It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months. In the case of water demand, we analyze the EDDI. It shows a positive trend, caused by an increase in AED. The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI). The behavior is similar to SPI, having close trends with opposite signs. Nevertheless, the spatial patterns are different. The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."

@@

The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale). Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI. Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.

@@

The vegetation productivity shows a high negative trend in the "norte chico" and "zona central" of continental Chile. This trend is in the primary case related to long-term precipitation decreases, which in the secondary case could be explained by the increase in AED. The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.

>>>>

---
title: "Assessment of drought in continental Chile for 1981–2023 by climate variables of water supply and demand, soil moisture, and vegetation"
format:
  ieee-pdf:
    keep-tex: false  
  ieee-html: default
title-block-banner: true
author:
  - id: dfolio
    name: Francisco Zambrano
    affiliations:
      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745
        orcid: 0000-0001-6896-8534
        email: francisco.zambrano@umayor.cl
        #url: https://dfolio.fr/
        attributes:
        corresponding: true
        #photo: david-folio.png
  - id: iduran
    name: Iongel Duran-Llacer
    affiliations: 
      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745

note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
#keywords: [IEEE, IEEEtran, journal, Quarto, Pandoc, template]
#funding: 
funding: 
  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022, Fondecyt de Iniciacipon 11190360 and Fondecyt de Postdoctorado 3230678"
pageheader:
  left: IGARSS, January 2024
  right: 'D. Folio:  A Sample Article Using quarto-ieee'
bibliography: references.bib
date: 2024-01-05
---

@@

A persistent drought is impacting Chile. It affects the hydrological system and vegetation development. Research studies have focused on the central part of the country. This is due to a persistent period of water scarcity. This scarcity has been found to be a megadrought. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December. The SPI only considers precipitation as a drought indicator. It does not account for atmospheric evaporative demand (AED), soil moisture, or their combined vegetation productivity. We present a developing database (DDS4Chl) of drought indices for continental Chile since 1981. The indices measure water demand, supply and its impact on vegetation. We derived the SPI for water demand. We also derived the Evaporative Demand Drought Index (EDDI) for water supply. The Standardized Precipitation Evapotranspiration Index (SPEI) shows the combined effect of AED and precipitation. We estimate the standardized anomaly of cumulative soil moisture at one meter (zcSM) and the standardized anomaly of cumulative NDVI (zcNDVI) to show the impact of water demand and supply. We present the historical linear trend of the drought indices in continental Chile. Also we calculated the temporal correlation between the indices of water supply and demand, and soil moisture with the zcNDVI (proxy of vegetation productivity). The dought index for soil moisture acummulated at 12 months (zcSM-12) showed to be the best predictor of vegetation productivity around continental Chile.
<br>

**Index Terms—** drought, Chile, water supply, water demand, soil moisture

@@

We evaluated the trend of the drought indices SPI, SPEI, EDDI, and zcSM; and their relation to vegetation productivity. Our goal is to assess the impact on vegetation of the climate variables of water demand and supply and soil moisture.

```{r fig-studyArea}
#| fig-pos: "!ht"
#| fig-env: "figure*"
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) Land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"

@@

Continental Chile has at least seven major climatic types. They range from low desert in the north to alpine tundra and glaciers in the east and southeast. The south has an Oceanic climate and central Chile has a Mediterranean climate [@Beck2023]. For the analysis, continental Chile was also divided into five zones according to a latitudinal gradient: "Norte Grande", "Norte Chico",  "Zona Central", "Zona Sur", and "Zona Asutral". (@fig-studyArea)

```{r fig-zcNDVI_var, out.extra='style="float:right; padding:10px"'}
#| fig-pos: "!ht"
#| fig-env: "figure*"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Zona Asutral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
```

@@

Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("Norte Chico" and "Zona Central") (@fig-zcNDVI_var). Indicating either a reduction of vegetation surface, biomass, or status. It reached its lowest values during the year 2020, mostly on the "Zona Central." There are other negative trends in small areas in "Norte Grande" and "zona asutral."

Analyzing the water supply in the macrozones "Norte Chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation. It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months. In the case of water demand, we analyze the EDDI. It shows a positive trend, caused by an increase in AED. The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "Norte Grande" and "Norte Chico" (@fig-trendEDDI). The behavior is similar to SPI, having close trends with opposite signs. Nevertheless, the spatial patterns are different. The maximum trend for SPI-36 is in "Norte Chico" and "Zona Central." For EDDI-36, the maximum trend is from "northe grande" to "Zona Central."

@@

The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale). Pearson correlations between 0.6 and 0.9 are concentrated in the "Norte Chico" and "Zona Central" for SPI and SPEI. Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI (@fig-corPerson).

@@

The vegetation productivity shows a high negative trend in the "Norte Chico" and "Zona Central" of continental Chile. This trend is in the primary case related to long-term precipitation decreases, which in the secondary case could be explained by the increase in AED. The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.
